---
title: "Uploading Data and Formatting"
author: "Anna Stouffer"
date: "2024-01-11"
output: html_document
---

# Andrew - this is the most chaotic file of the code >.<
# this uploads the death and temperature data, puts them in long formats, and combines them in different ways. The issue is that i basically continuously uploaded this file as I worked on my thesis, so as you get further down you have to run other scripts to generate the data that you then use. So basically like try to run this with the data you generated, and wherever it stops working bc you don't have the correct file (should happen when you need matched data), go to file #5

# you will need to run file #9 if you want to do any of the SVI stuff



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
################################################################################
# uploading data 
################################################################################

# all deaths (each row represents one death)
#death_data_all = read.csv("~/Thesis Work/Thesis Files/all_death_data.csv")
#death_data_all_stimi = read.csv("~/Thesis Work/Thesis Files/all_death_data_stimi_labeled.csv")
#death_data_labeled = read.csv("~/Thesis Work/Thesis Files/all_death_data_labeled.csv")
death_data_through_2022 = read.csv("~/Thesis Work/Thesis Files/all_death_data_2022.csv")

death_data_all = death_data_through_2022

# wide-format temperature data (each row is a county and each column is a day)
max_temperature_data = read.csv("~/Thesis Work/Thesis Files/max_temperature_per_county_per_day_20002022.csv")
min_temperature_data = read.csv("~/Thesis Work/Thesis Files/min_temperature_per_county_per_day_20002022.csv")
mean_temperature_data = read.csv("~/Thesis Work/Thesis Files/mean_temperature_per_county_per_day_20002022.csv")
vpdmin_data = read.csv("~/Thesis Work/Thesis Files/vpdmin_per_county_per_day_20002022.csv")
vpdmax_data = read.csv("~/Thesis Work/Thesis Files/vpdmax_per_county_per_day_20002022.csv")

vpdmin_data = vpdmax_data

# long-format county populations 
county_pop = read.csv("~/Thesis Work/Thesis Files/population_per_county_per_year.csv")

#SVI data 
SVI_data = read.csv("~/Thesis Work/Thesis Files/SVI_data.csv")


################################################################################
################################################################################
```

```{r}
# update this
# note on proportion of opioid deaths

# there are 1927190 total drug-related deaths in NC between 1999 and 2021
total_death_count = 1927190

# the following is the number of opioid-related deaths
dim(death_data_all)

print(paste("opioid deaths make up", round((dim(death_data_all)[1]/total_death_count)*100,2), "% of the total deaths in North Carolina over this time period"))

```

```{r}
# formatting the data appropriately
death_data_all$Date = as.Date(death_data_all$deathdate)


# reshaping the temperature data into a long format
max_temperature_data_long = max_temperature_data %>%
  gather(Date, Temperature, starts_with("X"), -GEOID, -county) %>%
  rename(county = "county")
min_temperature_data_long = min_temperature_data %>%
  gather(Date, Temperature, starts_with("X"), -GEOID, -county) %>%
  rename(county = "county")
mean_temperature_data_long = mean_temperature_data %>%
  gather(Date, Temperature, starts_with("X"), -GEOID, -county) %>%
  rename(county = "county")
vpdmin_data_long = vpdmin_data %>%
  gather(Date, MinVPDeficit, starts_with("X"), -GEOID, -county) %>%
  rename(county = "county")

# converting the date colunm to a date type
max_temperature_data_long$Date = as.Date(sub("X", "", max_temperature_data_long$Date), format = "%Y.%m.%d")
min_temperature_data_long$Date = as.Date(sub("X", "", min_temperature_data_long$Date), format = "%Y.%m.%d")
mean_temperature_data_long$Date = as.Date(sub("X", "", mean_temperature_data_long$Date), format = "%Y.%m.%d")
vpdmin_data_long$Date = as.Date(sub("X", "", vpdmin_data_long$Date), format = "%Y.%m.%d")

# adding a year column 
max_temperature_data_long$year = year(max_temperature_data_long$Date)
min_temperature_data_long$year = year(min_temperature_data_long$Date)
mean_temperature_data_long$year = year(mean_temperature_data_long$Date)
vpdmin_data_long$year = year(vpdmin_data_long$Date)

# renaming the fips column to GEOID from population data
county_pop = county_pop %>%
  rename(GEOID = fips)

max_temperature_data_long = max_temperature_data_long %>%
  left_join(county_pop, by = c("year", "county", "GEOID"))
min_temperature_data_long = min_temperature_data_long %>%
  left_join(county_pop, by = c("year", "county", "GEOID"))
mean_temperature_data_long = mean_temperature_data_long %>%
  left_join(county_pop, by = c("year", "county", "GEOID"))
vpdmin_data_long = vpdmin_data_long %>%
  left_join(county_pop, by = c("year", "county", "GEOID"))

# Convert date columns to date class 
max_temperature_data_long$Date = as.Date(max_temperature_data_long$Date)
min_temperature_data_long$Date = as.Date(min_temperature_data_long$Date)
mean_temperature_data_long$Date = as.Date(mean_temperature_data_long$Date)
vpdmin_data_long$Date = as.Date(vpdmin_data_long$Date)

# rename temperature column to more specific names
max_temperature_data_long <- max_temperature_data_long %>%
  rename(MaxTemperature = Temperature)
min_temperature_data_long <- min_temperature_data_long %>%
  rename(MinTemperature = Temperature)
mean_temperature_data_long <- mean_temperature_data_long %>%
  rename(MeanTemperature = Temperature)

death_data_all$Date = as.Date(death_data_all$deathdate)

# creating a column that has the death count for each date
combined_data = death_data_all %>%
  group_by(county, Date) %>%
  mutate(death_count = n()) %>%
  ungroup()


# selecting only the columns that I want
combined_data = combined_data %>%
  dplyr::select(county, Date, death_count, GEOID)

# combining the temperate data with the death data. Should be a row for each date
combined_data = max_temperature_data_long %>%
  left_join(combined_data, by = c("county", "Date", "GEOID"))
combined_data <- left_join(combined_data, min_temperature_data_long %>% dplyr::select(county, Date, GEOID, MinTemperature), by = c("county", "Date", "GEOID"))
combined_data <- left_join(combined_data, mean_temperature_data_long %>% dplyr::select(county, Date, GEOID, MeanTemperature), by = c("county", "Date", "GEOID"))
combined_data <- left_join(combined_data, vpdmin_data_long %>% dplyr::select(county, Date, GEOID, MinVPDeficit), by = c("county", "Date", "GEOID"))


# adding a death_count column
combined_data = combined_data %>%
  mutate(death_count = ifelse(is.na(death_count), 0, death_count))

# adding a death_rate column
combined_data$death_rate = combined_data$death_count/combined_data$total_population

# replacing NAs with 0s
combined_data = combined_data %>%
  mutate(death_rate = ifelse(is.na(death_rate), 0, death_rate))

head(combined_data)

```

```{r}
# removing duplicated in combined_data

dim(combined_data)

# removing duplicates from combined data
duplicates = combined_data %>% 
  group_by(county, Date) %>% 
  summarise(count = n()) %>% 
  filter(count > 1)

combined_data = combined_data %>% distinct(county, Date, .keep_all = TRUE)

dim(combined_data)
head(combined_data)

```

```{r}
# adding heat index 
#vpdmax in units hPa

# calculating saturation vapor pressure (units Pa)

# combined_data$SatVapPres = (exp(34.494-((4924.99)/(combined_data$MaxTemperature+273.1)))/((combined_data$MaxTemperature+105)^1.57))
# 
# (exp(43.494-((6545.8/(combined_data$MaxTemperature+278)))/((combined_data$MaxTemperature+868)^2))


# this didnt save and I need to check on it again
combined_data <- combined_data %>%
  mutate(
    SatVapPres = ifelse(MaxTemperature > 0,
                        exp(34.494 - (4924.99 / (MaxTemperature + 237.1))) / ((MaxTemperature + 105) ^ 1.57),
                        exp(43.494 - (6545.8 / (MaxTemperature + 278))) / ((MaxTemperature + 868) ^ 2))
  )

# converting VPDdeficiet from hPa to Pa
combined_data$MinVPDeficit = combined_data$MinVPDeficit*100

#vpdmax is actual vapor pressure - sat vapor pressure 
combined_data$VaporPressure =  combined_data$SatVapPres - combined_data$MinVPDeficit

# relative humidity is vp divided by saturation vapor pressure (multiplied by 100 because numerator is hPa and denom is Pa)

combined_data$RelHum = ((combined_data$VaporPressure)/combined_data$SatVapPres)*100

combined_data$MaxTemperatureF = (combined_data$MaxTemperature*(9/5)) + 32

combined_data$HeatIndex = heat.index(t = combined_data$MaxTemperatureF,
                                    rh = combined_data$RelHum,
                                    temperature.metric = "fahrenheit")



combined_data <- dplyr::select(combined_data, -SatVapPres, -MaxTemperatureF, -MinVPDeficit)

Heat_Index = combined_data %>%
  select(c("GEOID", "Date", "HeatIndex"))

# #Assuming combined_data is your data frame
# ggplot(combined_data, aes(x = Date, y = HeatIndex)) +
#   geom_point() +  # Plotting points
#   labs(x = "Date", y = "MaxVPDeficit") +  # Labeling axes
#   theme_minimal()  # Applying minimal theme


```


```{r}

# adding temperature to the death data 

# Left join and select only the Temperature column
death_data_all = left_join(death_data_all, combined_data %>% dplyr::select(county, Date, MaxTemperature, MinTemperature, MeanTemperature, HeatIndex, RelHum), by = c("county", "Date"), )

dim(death_data_all)
```


```{r}
# adding lagged temperature to data 

# Assuming 'Date' is in Date format
combined_data$Date <- as.Date(combined_data$Date)

# Calculate lagged temperature for the current date and two days before
combined_data <- combined_data %>%
  arrange(county, Date) %>%
  group_by(county) %>%
  mutate(
    max_lag_temp_0_2 = (MaxTemperature + lag(MaxTemperature, 1) + lag(MaxTemperature, 2)) / 3,
    min_lag_temp_0_2 = (MinTemperature + lag(MinTemperature, 1) + lag(MinTemperature, 2)) / 3,
    mean_lag_temp_0_2 = (MeanTemperature + lag(MeanTemperature, 1) + lag(MeanTemperature, 2)) / 3,
    max_lag_temp_0_5 = (MaxTemperature + lag(MaxTemperature, 1) + lag(MaxTemperature, 2) + lag(MaxTemperature, 3) + lag(MaxTemperature, 4) + lag(MaxTemperature, 5)) / 6,
    min_lag_temp_0_5 = (MinTemperature + lag(MinTemperature, 1) + lag(MinTemperature, 2) + lag(MinTemperature, 3) + lag(MinTemperature, 4) + lag(MinTemperature, 5)) / 6,
    mean_lag_temp_0_5 = (MeanTemperature + lag(MeanTemperature, 1) + lag(MeanTemperature, 2) + lag(MeanTemperature, 3) + lag(MeanTemperature, 4) + lag(MeanTemperature, 5)) / 6,
  ) %>%
  ungroup()

head(combined_data)
# Left join and select relevant columns
#death_data_all <- left_join(death_data_all, combined_data %>% select(county, Date, lag_temp_0_2), by = c("county", "Date"))

```


```{r}
# organizing the data in different ways

# creating a binary death variable 
binary_death_data = combined_data %>%
  mutate(binary_death = ifelse(death_count == 0, 0, 1))

# aggregating data for the whole state
data_aggregate = combined_data %>%
  group_by(Date) %>%
  summarise(
    MaxTemperature = mean(MaxTemperature),
    MinTemperature = mean(MinTemperature),
    MeanTemperature = mean(MeanTemperature),
    HeatIndex = mean(HeatIndex),
    RelHum = mean(RelHum),
    Total_Death_Count = sum(death_count),
    Population = sum(total_population)
  )

# Create a new column 'simple_race' considering ethnicity
death_data_all <- death_data_all %>%
  mutate(
    simple_race = case_when(
      grepl("White", race) & !grepl("Other", race) ~ "White Non-Hispanic",
      grepl("Black", race) & ethnicity == "N" ~ "Black Non-Hispanic",
      grepl("Asian|Japanese|Chinese|Filipino", race) & ethnicity == "N" ~ "Asian Non-Hispanic",
      grepl("American Indian|Native Hawaiian", race) & ethnicity == "N" ~ "Native American Non-Hispanic",
      ethnicity != "N" ~ "Any Race - Hispanic",
      TRUE ~ "Other"
    )
  )

head(death_data_all)

# Create a new column 'simple_ethnicity'
death_data_all <- death_data_all %>%
  mutate(simple_ethnicity = ifelse(ethnicity == "N", "No", "Hispanic"))

# creating age brackets 
death_data_all = death_data_all %>%
  mutate(age = as.numeric(age),
         simple_age = cut(age, breaks = c(-Inf, 17, 24, 34, 44, 54, Inf),
                         labels = c("0-17", "18-24", "25-34", "35-44", "45-54", "55+"),
                         include.lowest = TRUE))

```

```{r}
# adding a column that assigns warm season, cold season, or shoulder season category

head(death_data_all)

# Create a vector to store the seasons
season_vector <- character(nrow(death_data_all))

# Assign seasons based on the criteria
season_vector[months(death_data_all$Date) %in% c("May", "June", "July", "August", "September", "October")] <- "warm"
season_vector[months(death_data_all$Date) %in% c("December", "January", "February", "March")] <- "cool"
season_vector[months(death_data_all$Date) %in% c("April", "November")] <- "shoulder"

# Add the season column to the dataframe
death_data_all$season <- season_vector

```

```{r}
# assigning blocks to the data 

# setting as warm and cool season
season_vector <- character(nrow(combined_data))

# Assign seasons based on the criteria
season_vector[months(combined_data$Date) %in% c("May", "June", "July", "August", "September", "October")] <- "warm"
season_vector[months(combined_data$Date) %in% c("December", "January", "February", "March")] <- "cool"
season_vector[months(combined_data$Date) %in% c("April", "November")] <- "shoulder"

# Add the season column to the dataframe
combined_data$season <- season_vector
```

```{r}
# making a warm season dataframe
data_warm_season <- combined_data %>%
  filter(season == "warm")
# making blocks in the warm season datafram
data_warm_season <- data_warm_season %>%
  mutate(
    block = (as.numeric(format(Date, "%j")) - as.numeric(format(as.Date(paste0(format(Date, "%Y"), "-05-01")), "%j"))) %/% 21 + 1
  )

# making a cool season dataframe
data_cool_season <- combined_data %>%
  filter(season == "cool")
data_cool_season <- data_cool_season %>%
  mutate(
    month_day = format(Date, "%m-%d"),
    block = case_when(
      month_day %in% c("12-01", "12-02", "12-03", "12-04", "12-05", "12-06", "12-07", "12-08", "12-09", "12-10", "12-11", "12-12", "12-13", "12-14", "12-15", "12-16", "12-17", "12-18", "12-19", "12-20", "12-21") ~ 1,
      month_day %in% c("12-22", "12-23", "12-24", "12-25", "12-26", "12-27", "12-28", "12-29", "12-30", "12-31", "01-01", "01-02", "01-03", "01-04", "01-05", "01-06", "01-07", "01-08", "01-09", "01-10", "01-11") ~ 2,
      month_day %in% c("01-12", "01-13", "01-14", "01-15", "01-16", "01-17", "01-18", "01-19", "01-20", "01-21", "01-22", "01-23", "01-24", "01-25", "01-26", "01-27", "01-28", "01-29", "01-30", "01-31", "02-01") ~ 3,
      month_day %in% c("02-02", "02-03", "02-04", "02-05", "02-06", "02-07", "02-08", "02-09", "02-10", "02-11", "02-12", "02-13", "02-14", "02-15", "02-16", "02-17", "02-18", "02-19", "02-20", "02-21") ~ 4,
      month_day %in% c("02-22", "02-23", "02-24", "02-25", "02-26", "02-27", "02-28", "02-29", "03-01", "03-02", "03-03", "03-04", "03-05", "03-06", "03-07", "03-08", "03-09", "03-10", "03-11", "03-12", "03-13", "03-14") ~ 5,
      TRUE ~ 6  # Default case, assign a block for any remaining dates
      # Add more cases if needed
    )
  )

data_cool_season <- dplyr::select(data_cool_season, -month_day)

head(data_cool_season, 1000)


```


```{r}
# will remove this once i run my data matching function 
# 
# matched_data = read.csv("~/Thesis Work/Thesis Files/matched_data.csv")
# 
# # adding the uniqueID columns for all the data
# matched_data <- matched_data %>%
#   group_by(county) %>%
#   mutate(uniqueID = ifelse(is.na(uniqueID), lag(uniqueID), uniqueID)) %>%
#   fill(uniqueID, .direction = "down") %>%
#   ungroup()
# 
# # filling in the NAs in the "year" and "total population" columns
# matched_data <- matched_data %>%
#   group_by(county) %>%
#   mutate(
#     year = ifelse(is.na(year), lead(year), year),
#     total_population = ifelse(is.na(total_population), lead(total_population), total_population)
#   ) %>%
#   fill(c("year", "total_population"), .direction = "down") %>%
#   ungroup()
# 
# head(matched_data)
# 
# # add the sex, race, ethnicity, and age columns to the case-control data
# matched_data <- left_join(matched_data, death_data_all %>% select(uniqueID, simple_race, simple_age, simple_ethnicity, sex), by = "uniqueID")
# 
# # adding lag temperature column
# matched_data$Date = as.Date(matched_data$Date)
# matched_data <- left_join(matched_data, combined_data %>% select(county, Date, max_lag_temp_0_2, min_lag_temp_0_2, mean_lag_temp_0_2, max_lag_temp_0_5, min_lag_temp_0_5, mean_lag_temp_0_5), by = c("county", "Date"))
# 
# matched_data = na.omit(matched_data)
# 
# # Create a vector to store the seasons
# season_vector <- character(nrow(matched_data))
# 
# matched_data$Date = as.Date(matched_data$Date)
# 
# # Assign seasons based on the criteria
# season_vector[months(matched_data$Date) %in% c("May","June", "July", "August", "September", "October")] <- "warm"
# season_vector[months(matched_data$Date) %in% c("December", "January", "February", "March")] <- "cool"
# season_vector[months(matched_data$Date) %in% c("April", "November")] <- "shoulder"
# 
# # Add the season column to the dataframe
# matched_data$season <- season_vector
# 
# head(matched_data)
# 

```

```{r}
# # only warm season deaths
warm_season_deaths = death_data_all %>%
  filter(season == "warm")

head(warm_season_deaths)

# assigning blocks to warm season deaths
warm_season_deaths <- warm_season_deaths %>%
  mutate(
    block = (as.numeric(format(Date, "%j")) - as.numeric(format(as.Date(paste0(format(Date, "%Y"), "-05-01")), "%j"))) %/% 21 + 1
  )

# warm_season_deaths_long_sept = warm_season_deaths_long_sept %>%
#   filter(block %in% c(1, 2, 3, 4, 5, 6, 7, 8))
# 
# # assigning blocks to warm season deaths
# warm_season_deaths <- warm_season_deaths %>%
#   mutate(
#     block = (as.numeric(format(Date, "%j")) - as.numeric(format(as.Date(paste0(format(Date, "%Y"), "-05-01")), "%j"))) %/% 21 + 1
#   )
# 
# #dim(warm_season_deaths_short_sept)
# 
# 
# only cool season deaths

cool_season_deaths = death_data_all %>%
  filter(season == "cool")
 
cool_season_deaths$Date = as.Date(cool_season_deaths$Date)

cool_season_deaths <- cool_season_deaths %>%
  mutate(
    month_day = format(Date, "%m-%d"),
    block = case_when(
      month_day %in% c("12-01", "12-02", "12-03", "12-04", "12-05", "12-06", "12-07", "12-08", "12-09", "12-10", "12-11", "12-12", "12-13", "12-14", "12-15", "12-16", "12-17", "12-18", "12-19", "12-20", "12-21") ~ 1,
      month_day %in% c("12-22", "12-23", "12-24", "12-25", "12-26", "12-27", "12-28", "12-29", "12-30", "12-31", "01-01", "01-02", "01-03", "01-04", "01-05", "01-06", "01-07", "01-08", "01-09", "01-10", "01-11") ~ 2,
      month_day %in% c("01-12", "01-13", "01-14", "01-15", "01-16", "01-17", "01-18", "01-19", "01-20", "01-21", "01-22", "01-23", "01-24", "01-25", "01-26", "01-27", "01-28", "01-29", "01-30", "01-31", "02-01") ~ 3,
      month_day %in% c("02-02", "02-03", "02-04", "02-05", "02-06", "02-07", "02-08", "02-09", "02-10", "02-11", "02-12", "02-13", "02-14", "02-15", "02-16", "02-17", "02-18", "02-19", "02-20", "02-21") ~ 4,
      month_day %in% c("02-22", "02-23", "02-24", "02-25", "02-26", "02-27", "02-28", "02-29", "3-01", "3-02", "3-03", "3-04", "3-05", "3-06", "3-07", "3-08", "3-09", "3-10", "3-11", "3-12", "3-13", "3-14") ~ 5,
      TRUE ~ 6  # Default case, assign a block for any remaining dates
      # Add more cases if needed
    )
  )

cool_season_deaths <- dplyr::select(cool_season_deaths, -month_day)

# 
 head(cool_season_deaths, 1000)

```

```{r}
# adjusting warm-season matched data

warm_season_matched_data_blocks = read.csv("~/Thesis Work/Thesis Files/warm_season_matched_data_blocks_2022.csv")

head(warm_season_matched_data_blocks)

warm_season_matched_data_blocks <- warm_season_matched_data_blocks[, !names(warm_season_matched_data_blocks) %in% "HeatIndex"]

Heat_Index$Date = as.Date(Heat_Index$Date)
warm_season_matched_data_blocks$Date = as.Date(warm_season_matched_data_blocks$Date)

merged_data <- left_join(warm_season_matched_data_blocks, Heat_Index, by = c("GEOID", "Date"))

# Replace HeatIndex in warm_season_matched_data_blocks with HeatIndex from combined_data
warm_season_matched_data_blocks$HeatIndex <- merged_data$HeatIndex

# Assuming 'warm_season_matched_data_blocks' is your dataframe
warm_season_matched_data_blocks <- warm_season_matched_data_blocks %>%
  mutate(
    binary = ifelse(case_control == "case", 1, 0)
  )

# # adding lag temperature column
# warm_season_matched_data_blocks$Date = as.Date(warm_season_matched_data_blocks$Date)
# warm_season_matched_data_blocks <- left_join(warm_season_matched_data_blocks, combined_data %>% select(county, Date, max_lag_temp_0_2, min_lag_temp_0_2, mean_lag_temp_0_2, max_lag_temp_0_5, min_lag_temp_0_5, mean_lag_temp_0_5), by = c("county", "Date"))

warm_season_matched_data_blocks$Date = as.Date(warm_season_matched_data_blocks$Date)

# # adding min, mean, and max temperature column (can delete this if I rerun the function for matching data)
# warm_season_matched_data_blocks <- left_join(warm_season_matched_data_blocks, combined_data %>% select(county, Date, MinTemperature, MaxTemperature, MeanTemperature, HeatIndex), by = c("county", "Date"))
# head(warm_season_matched_data_blocks)
# warm_season_matched_data_blocks <- select(warm_season_matched_data_blocks, -Temperature)


# # fixing race column
# warm_season_matched_data_blocks <- warm_season_matched_data_blocks %>%
#   mutate(
#     simple_race = case_when(
#       grepl("White", simple_race) & ethnicity == "No" ~ "White Non-Hispanic",
#       grepl("Black", simple_race) & ethnicity == "No" ~ "Black Non-Hispanic",
#       grepl("Asian|Japanese|Chinese|Filipino", simple_race) & ethnicity == "No" ~ "Asian Non-Hispanic",
#       grepl("American Indian|Native Hawaiian|Native American", simple_race) & ethnicity == "No" ~ "Native American Non-Hispanic",
#       ethnicity == "Hispanic" ~ "Any Race - Hispanic",
#       TRUE ~ "Other"
#     )
#   )

unique(warm_season_matched_data_blocks$simple_race)

head(warm_season_matched_data_blocks, 1000)

```

```{r}

# adjusting cool-season matched data

cool_season_matched_data_blocks = read.csv("~/Thesis Work/Thesis Files/cool_season_matched_data_blocks_2022.csv")

cool_season_matched_data_blocks <- cool_season_matched_data_blocks[, !names(cool_season_matched_data_blocks) %in% "HeatIndex"]

cool_season_matched_data_blocks$Date = as.Date(cool_season_matched_data_blocks$Date)

merged_data <- left_join(cool_season_matched_data_blocks, Heat_Index, by = c("GEOID", "Date"))

# Replace HeatIndex in warm_season_matched_data_blocks with HeatIndex from combined_data
cool_season_matched_data_blocks$HeatIndex <- merged_data$HeatIndex

# Assuming 'warm_season_matched_data_blocks' is your dataframe
cool_season_matched_data_blocks <- cool_season_matched_data_blocks %>%
  mutate(
    binary = ifelse(case_control == "case", 1, 0)
  )

# adding lag temperature column
cool_season_matched_data_blocks$Date = as.Date(cool_season_matched_data_blocks$Date)
cool_season_matched_data_blocks <- left_join(cool_season_matched_data_blocks, combined_data %>% dplyr::select(county, Date, max_lag_temp_0_2, min_lag_temp_0_2, mean_lag_temp_0_2, max_lag_temp_0_5, min_lag_temp_0_5, mean_lag_temp_0_5), by = c("county", "Date"))

# fixing race column
cool_season_matched_data_blocks <- cool_season_matched_data_blocks %>%
  mutate(
    simple_race = case_when(
      grepl("White", simple_race) & ethnicity == "No" ~ "White Non-Hispanic",
      grepl("Black", simple_race) & ethnicity == "No" ~ "Black Non-Hispanic",
      grepl("Asian|Japanese|Chinese|Filipino", simple_race) & ethnicity == "No" ~ "Asian Non-Hispanic",
      grepl("American Indian|Native Hawaiian|Native American", simple_race) & ethnicity == "No" ~ "Native American Non-Hispanic",
      ethnicity == "Hispanic" ~ "Any Race - Hispanic",
      TRUE ~ "Other"
    )
  )

```



```{r}
# # adjusting the block periods for analytics 
# 
# cool_season_deaths = death_data_all %>%
#   filter(season == "cool")
# 
# cool_season_deaths$Date = as.Date(cool_season_deaths$Date)
# 
# cool_season_deaths_short_feb <- cool_season_deaths %>%
#   mutate(
#     month_day = format(Date, "%m-%d"),
#     block = case_when(
#       month_day %in% c("12-01", "12-02", "12-03", "12-04", "12-05", "12-06", "12-07", "12-08", "12-09", "12-10", "12-11", "12-12", "12-13", "12-14", "12-15", "12-16", "12-17", "12-18", "12-19", "12-20", "12-21") ~ 1,
#       month_day %in% c("12-22", "12-23", "12-24", "12-25", "12-26", "12-27", "12-28", "12-29", "12-30", "12-31", "01-01", "01-02", "01-03", "01-04", "01-05", "01-06", "01-07", "01-08", "01-09", "01-10", "01-11") ~ 2,
#       month_day %in% c("01-12", "01-13", "01-14", "01-15", "01-16", "01-17", "01-18", "01-19", "01-20", "01-21", "01-22", "01-23", "01-24", "01-25", "01-26", "01-27", "01-28", "01-29", "01-30", "01-31", "02-01") ~ 3,
#       month_day %in% c("02-02", "02-03", "02-04", "02-05", "02-06", "02-07", "02-08", "02-09", "02-10", "02-11", "02-12", "02-13", "02-14", "02-15", "02-16", "02-17", "02-18", "02-19", "02-20", "02-21") ~ 4,
#       TRUE ~ 6  # Default case, assign a block for any remaining dates
#       # Add more cases if needed
#     )
#   )
# 
# cool_season_deaths_short_feb <- select(cool_season_deaths_short_feb, -month_day)
# 
# cool_season_deaths_long_feb <- cool_season_deaths %>%
#   mutate(
#     month_day = format(Date, "%m-%d"),
#     block = case_when(
#       month_day %in% c("12-01", "12-02", "12-03", "12-04", "12-05", "12-06", "12-07", "12-08", "12-09", "12-10", "12-11", "12-12", "12-13", "12-14", "12-15", "12-16", "12-17", "12-18", "12-19", "12-20", "12-21") ~ 1,
#       month_day %in% c("12-22", "12-23", "12-24", "12-25", "12-26", "12-27", "12-28", "12-29", "12-30", "12-31", "01-01", "01-02", "01-03", "01-04", "01-05", "01-06", "01-07", "01-08", "01-09", "01-10", "01-11") ~ 2,
#       month_day %in% c("01-12", "01-13", "01-14", "01-15", "01-16", "01-17", "01-18", "01-19", "01-20", "01-21", "01-22", "01-23", "01-24", "01-25", "01-26", "01-27", "01-28", "01-29", "01-30", "01-31", "02-01") ~ 3,
#       month_day %in% c("02-02", "02-03", "02-04", "02-05", "02-06", "02-07", "02-08", "02-09", "02-10", "02-11", "02-12", "02-13", "02-14", "02-15", "02-16", "02-17", "02-18", "02-19", "02-20", "02-21") ~ 4,
#       month_day %in% c("02-22", "02-23", "02-24", "02-25", "02-26", "02-27", "02-28", "02-29", "3-01", "3-02", "3-03", "3-04", "3-05", "3-06", "3-07", "3-08", "3-09", "3-10", "3-11", "3-12", "3-13", "3-14") ~ 5,
# 
#       TRUE ~ 6  # Default case, assign a block for any remaining dates
#       # Add more cases if needed
#     )
#   )
# 
# cool_season_deaths_long_feb <- select(cool_season_deaths_long_feb, -month_day)
# 
# 
# warm_season_deaths_short_sept = warm_season_deaths %>%
#   filter(block %in% c(1, 2, 3, 4, 5, 6, 7))
# 
# dim(warm_season_deaths_short_sept)
# 
# head(warm_season_matched_data_blocks)
# 
# head(cool_season_deaths_short_feb)

```


```{r}
# 
# matched_data_warm_blocks = read.csv("~/Thesis Work/Thesis Files/warm_season_matched_blocks_updated_ICD10.csv")
# head(matched_data_warm_blocks, 1000)
# names(combined_data)
# 
# # Merge matched_data_warm_blocks with combined_data to add HeatIndex column
# matched_data_warm_blocks <- merge(matched_data_warm_blocks, 
#                                    combined_data[c("county", "Date", "HeatIndex")],
#                                    by = c("county", "Date"),
#                                    all.x = TRUE)
# 
# # View the updated dataset
# head(matched_data_warm_blocks)



```

```{r}
# adding SVI tertile to data

death_data_all <- left_join(death_data_all, SVI_data, by = "GEOID")

warm_season_matched_data_blocks = left_join(warm_season_matched_data_blocks, SVI_data, by = "GEOID")
combined_data

warm_season_matched_data_blocks <- merge(warm_season_matched_data_blocks, combined_data[, c("Date", "county", "RelHum")], by = c("Date", "county"), all.x = TRUE)


cool_season_matched_data_blocks = left_join(cool_season_matched_data_blocks, SVI_data, by = "GEOID")

cool_season_matched_data_blocks <- merge(cool_season_matched_data_blocks, combined_data[, c("Date", "county", "RelHum")], by = c("Date", "county"), all.x = TRUE)

# waiting for code from frank for this 


```


```{r}
# adding in race weighted temp for max temp
warm_season_matched_data_blocks

race_weighted_max_temp = read.csv("~/Thesis Work/Thesis Files/RaceWeightedMaxTemp.csv")

race_weighted_max_temp <- race_weighted_max_temp %>%
  rename(GEOID = County)

warm_season_matched_data_blocks$Date = as.Date(warm_season_matched_data_blocks$Date)
race_weighted_max_temp$Date = as.Date(race_weighted_max_temp$Date)

warm_season_matched_data_blocks <- warm_season_matched_data_blocks %>%
  left_join(race_weighted_max_temp, by = c("GEOID", "Date"))

warm_season_matched_data_blocks <- warm_season_matched_data_blocks %>%
  mutate(RaceWeightedMaxTemp = case_when(
    simple_race == "White Non-Hispanic" ~ WhiteMaxTemp,
    simple_race == "Black Non-Hispanic" ~ BlackMaxTemp,
    simple_race == "Asian Non-Hispanic" ~ AsianMaxTemp,
    simple_race == "Any Race - Hispanic" ~ HispanicMaxTemp,
    simple_race == "Native American Non-Hispanic" ~ MaxTemperature,
    simple_race == "Other" ~ MaxTemperature,
    TRUE ~ NA_real_
  ))

warm_season_matched_data_blocks <- warm_season_matched_data_blocks %>%
  select(-WhiteMaxTemp, -BlackMaxTemp, -AsianMaxTemp, -HispanicMaxTemp)

```

```{r}
# adding in race weighted temp for min temp
cool_season_matched_data_blocks

race_weighted_min_temp = read.csv("~/Thesis Work/Thesis Files/RaceWeightedMinTemp.csv")

race_weighted_min_temp <- race_weighted_min_temp %>%
  rename(GEOID = County)

cool_season_matched_data_blocks$Date = as.Date(cool_season_matched_data_blocks$Date)
race_weighted_min_temp$Date = as.Date(race_weighted_min_temp$Date)

cool_season_matched_data_blocks <- cool_season_matched_data_blocks %>%
  left_join(race_weighted_min_temp, by = c("GEOID", "Date"))

cool_season_matched_data_blocks <- cool_season_matched_data_blocks %>%
  mutate(RaceWeightedMinTemp = case_when(
    simple_race == "White Non-Hispanic" ~ WhiteMinTemp,
    simple_race == "Black Non-Hispanic" ~ BlackMinTemp,
    simple_race == "Asian Non-Hispanic" ~ AsianMinTemp,
    simple_race == "Any Race - Hispanic" ~ HispanicMinTemp,
    simple_race == "Native American Non-Hispanic" ~ MinTemperature,
    simple_race == "Other" ~ MinTemperature,
    TRUE ~ NA_real_
  ))

cool_season_matched_data_blocks <- cool_season_matched_data_blocks %>%
  select(-WhiteMinTemp, -BlackMinTemp, -AsianMinTemp, -HispanicMinTemp)

```

